---
title: "Subtype rare cns tumors using hallmark fusions"
output: html_notebook
author: "Jo Lynne Rokita"
date: 2026
---

In this PR, we identify subtype Rare CNS RNA-Seq biospecimens from `histologies-base.tsv` using hallmark fusions per [#115](https://github.com/rokitalab/OpenPedCan-Project-CNH/issues/115), removing RMS and EWS.

Fusions:
MN1::BEND2
MN1::CXXC5
BRD4::LEUTX
FUS::ETV4
FET::CREB --> this is really now CREB family, consisting of CREB1, ATF1, CREM partners

Genes:
PLAG
MN1
PLAGL
CXXC5
BCOR
PATZ1
CREB1
ATF1
CREM

The following subtypes are included in this step:

- Rare CNS, MN1::BEND2
- Rare CNS, BCOR-fused
- Rare CNS, BRD4::LEUTX
- Rare CNS, MN1::CXXC5
- Rare CNS, PATZ1-fused
- Rare CNS, PLAGL1-fused
- Rare CNS, FET::CREB

### Setup

```{r}
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "molecular-subtyping-Rare-CNS")
fus_input_dir <- file.path(root_dir, "analyses", "fusion-summary", "results")
subset_dir <- file.path(analysis_dir, "rare-cns-subset")
results_dir <- file.path(analysis_dir, "results")

if(!dir.exists(results_dir)){
  dir.create(results_dir)
}
```

## Input

# Candidate Rare CNS tumor specimens based on RNA-Seq fusions

We will use the fusion summary file to find samples that contain fusions defining rare CNS tumosr (e.g. _BCOR_ or _BCORL1_).
[These](https://pubmed.ncbi.nlm.nih.gov/40724977/) tumors are primarily driven by gene fusions involving the BCOR (BCL6 corepressor) or BCORL1 genes. Common fusion partners include EP300, CREBBP, and occasionally others.

```{r}
termslist <- jsonlite::fromJSON(
  file.path(subset_dir, "rare_cns_subtyping_path_dx_strings.json")
)

hist <- read_tsv(file.path (data_dir, "histologies-base.tsv"))
fus_mat <- read_tsv(file.path(fus_input_dir, "fusion_summary_rare_cns_foi.tsv"))

methyl_matches <- hist %>%
  filter(cohort == "PBTA",
         experimental_strategy == "Methylation",
         # let's remove EWS since they're subtyped in their own module
         pathology_diagnosis != "Ewings Sarcoma",
         # remove exclude dx
         !pathology_diagnosis %in% termslist$exclude_dx,
         !pathology_free_text_diagnosis %in% termslist$exclude_free,
         !sample_id %in% termslist$exclude_sample_ids,
         !is.na(pathology_diagnosis),
         # no DMGs
         !(dkfz_v12_methylation_subclass %in% termslist$exclude_methyl & dkfz_v12_methylation_subclass_score >=0.8)
  ) %>%
  pull(match_id) %>%
  unique()
        

cns_rna_biospecimens <- hist %>%
  filter(cohort == "PBTA",
         # let's remove EWS since they're subtyped in their own module
         pathology_diagnosis != "Ewings Sarcoma",
         # remove exclude dx
         !pathology_diagnosis %in% termslist$exclude_dx,
         !pathology_free_text_diagnosis %in% termslist$exclude_free,
         !sample_id %in% termslist$exclude_sample_ids,
         !is.na(pathology_diagnosis),
         !is.na(RNA_library)) 

# Which samples have these fusions?
fusion_hits <- fus_mat %>%
  filter(Kids_First_Biospecimen_ID %in% cns_rna_biospecimens$Kids_First_Biospecimen_ID) %>%
  filter(if_any(-Kids_First_Biospecimen_ID, ~ . > 0)) %>%
  pivot_longer(-Kids_First_Biospecimen_ID,
               names_to = "fusion",
               values_to = "value") %>%
  filter(value > 0) %>%
  select(-value) %>%
  # group if >1 fusion per bs id
  group_by(Kids_First_Biospecimen_ID) %>%
  summarise(fusion = paste(unique(fusion), collapse = ";"),
            .groups = "drop") %>%
  left_join(cns_rna_biospecimens[,c("Kids_First_Biospecimen_ID", "pathology_diagnosis", "pathology_free_text_diagnosis", "match_id")]) %>%
  write_tsv(file.path(results_dir, "rare_cns_fusion_hits.tsv"))
  
```
Subtype based on fusions

```{r subtype fusions}
fusion_subtypes <- fusion_hits %>%
  # split multi-fusion entries
  separate_rows(fusion, sep = ";") %>%

  # assign subtype using hybrid exact + pattern matching
  mutate(
    molecular_subtype = case_when(

      # exact hallmark fusions
      fusion == "MN1--BEND2" ~ "Rare CNS, MN1::BEND2",
      fusion == "MN1--CXXC5" ~ "Rare CNS, MN1::CXXC5",
      fusion == "BRD4--LEUTX" ~ "Rare CNS, BRD4::LEUTX",
      fusion %in% c("EP300--BCOR", "BCOR--EP300") ~ "Rare CNS, BCOR-fused",
      fusion %in% c("MN1--PATZ1", "EWSR1--PATZ1") ~ "Rare CNS, PATZ1-fused",
      fusion == "EWSR1--PLAGL1" ~ "Rare CNS, PLAGL1-fused",

      # CREB family beyond FET
      str_detect(fusion, "^(FUS|FET)--(CREB|ATF|CREM)") |
      str_detect(fusion, "(CREB1|ATF1|CREM)") ~ "Rare CNS, CREB-fused",
    
      # MN1 family members

      # BCOR family beyond EP300
      str_detect(fusion, "^BCOR|^BCORL1") ~ "Rare CNS, BCOR-fused",

      TRUE ~ NA_character_
    )
  ) %>%

  # collapse back to sample level
  group_by(Kids_First_Biospecimen_ID) %>%
  summarise(
    molecular_subtype = first(na.omit(molecular_subtype)),
    fusion = paste(unique(fusion), collapse = ";"),
    pathology_diagnosis = first(pathology_diagnosis),
    pathology_free_text_diagnosis = first(pathology_free_text_diagnosis),
    match_id = match_id,
    .groups = "drop"
  )

rna_methyl <- hist %>%
  filter(match_id %in% methyl_matches,
         dkfz_v12_methylation_subclass_score >=0.8) %>%
  select(match_id, dkfz_v12_methylation_subclass, dkfz_v12_methylation_subclass_score) %>%
  unique() %>%
  right_join(fusion_subtypes)
```
## Session Info

```{r}
sessionInfo()
```
