---
title: "Subset files for Rare CNS subtyping"
output: html_notebook
author: "Aylar Babaei, Jo Lynne Rokita"
date: 2026
---

In this PR, we identify candidate Rare CNS biospecimens from `histologies-base.tsv` and gather high-confidence methylation subtype calls for downstream Rare CNS subtyping.

As per issue [#114](https://github.com/rokitalab/OpenPedCan-Project-CNH/issues/114), Rare CNS subtyping is primarily based on high-confidence methylation subclassifications with `dkfz_v12_methylation_subclass_score >= 0.8`. Hallmark lesion evidence will be added in downstream scripts.

The following Rare CNS methylation classes are included in this step:

- Rare CNS, MN1::BEND2
- Rare CNS, ANTCON
- Rare CNS, ARMS
- Rare CNS, BCOR-fused
- Rare CNS, BCOR ITD
- Rare CNS, FOXR2 NBL
- Rare CNS, CIC SARC
- Rare CNS, DICER1 SARC
- Rare CNS, ERMS
- Rare CNS, BRD4::LEUTX
- Rare CNS, PLAG AMP
- Rare CNS, EWS
- Rare CNS, MYCN NBL
- Rare CNS, TMM NEG NBL
- Rare CNS, TMM POS NBL
- Rare CNS, MN1::CXXC5
- Rare CNS, PATZ1-fused
- Rare CNS, PLAGL1-fused
- Rare CNS, MYOD1 RMS

### Setup

```{r}
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

subset_dir <- file.path(
  root_dir,
  "analyses",
  "molecular-subtyping-Rare-CNS",
  "rare-cns-subset"
)

results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-Rare-CNS", "results")

if(!dir.exists(results_dir)){
  dir.create(results_dir)
}
```

## Input

```{r}
# File from 00-Rare-CNS-select-pathology-dx.R that is used for pathology
# diagnosis inclusion, exclusion criteria and Rare CNS methylation mapping
termslist <- jsonlite::fromJSON(
  file.path(subset_dir, "rare_cns_subtyping_path_dx_strings.json")
)

# subtype lookup table
subtype_to_opc_map <- tribble(
  ~NIH_v2_methylation_Class, ~dkfz_v12_methylation_subclass, ~OPC_molecular_subtype,
  "HGNET_MN1", "ABM_MN1", "Rare CNS, MN1::BEND2",
  "ANTCON", "ANTCON", "Rare CNS, ANTCON",
  "RMS_alevolar", "ARMS", "Rare CNS, ARMS",
  "EP300_BCOR", "CNS_BCOR_FUS", "Rare CNS, BCOR-fused",
  "HGNET_BCOR", "CNS_BCOR_ITD", "Rare CNS, BCOR ITD",
  "CNS_NB_FOXR2", "CNS_NB_FOXR2", "Rare CNS, FOXR2 NBL",
  "EFT_CIC", "CNS_SARC_CIC", "Rare CNS, CIC SARC",
  "CNS_SCS_DICER", "CNS_SARC_DICER", "Rare CNS, DICER1 SARC",
  "RMS_embryonal", "ERMS", "Rare CNS, ERMS",
  "ET_BRD4_LEUTX", "ET_BRD4_LEUTX", "Rare CNS, BRD4::LEUTX",
  "HGNET_PLAG", "ET_PLAG", "Rare CNS, PLAG AMP",
  "EWS", "EWS", "Rare CNS, EWS",
  "HGNET_CXXC5", "NET_CXXC5", "Rare CNS, MN1::CXXC5",
  "HGNET_PATZ", "NET_PATZ1", "Rare CNS, PATZ1-fused",
  "PLAGL1_FUS", "NET_PLAGL1_FUS", "Rare CNS, PLAGL1-fused",
  "RMS_MYOD1", "RMS_MYOD1", "Rare CNS, MYOD1 RMS",
  "CRINET", "CRINET", "Rare CNS, CRINET"
)

# Clinical file
hist <- readr::read_tsv(
  file.path(root_dir, "data", "histologies-base.tsv"),
  guess_max = 100000
)
```

## pathology-subset 

```{r}
# Candidate Rare CNS tumor specimens based on methylation
rare_cns_methyl <- hist %>%
  filter(experimental_strategy == "Methylation" & 
           (dkfz_v12_methylation_subclass_score >= 0.8 & 
              dkfz_v12_methylation_subclass %in% termslist$include_dkfz_subtypes) |
          (NIH_v2_methylation_Superfamily_mean_score >= 0.9 & 
             NIH_v2_methylation_Class_mean_score >= 0.9 & 
             NIH_v2_methylation_Class %in% termslist$include_nih_subtypes)
  ) %>%
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, match_id, sample_id, composition, 
         dkfz_v12_methylation_subclass, dkfz_v12_methylation_subclass_score,
         NIH_v2_methylation_Class, NIH_v2_methylation_Superfamily_mean_score, NIH_v2_methylation_Class_mean_score) %>%
  left_join(subtype_to_opc_map[,c("dkfz_v12_methylation_subclass", "OPC_molecular_subtype")]) %>%
  # assign DKFZ first
  mutate(molecular_subtype = case_when(
    dkfz_v12_methylation_subclass_score >= 0.8 ~ OPC_molecular_subtype,
    TRUE ~ NA_character_
  )) %>%
  # join NIH mapping
  left_join(subtype_to_opc_map[,c("NIH_v2_methylation_Class", "OPC_molecular_subtype")]) %>%
  # use NIH mapping where DKFZ missing
  mutate(molecular_subtype = case_when(
    is.na(molecular_subtype) ~ OPC_molecular_subtype,
    TRUE ~ molecular_subtype
  )) %>%
  # ensure consistent prefix
  mutate(molecular_subtype = if_else(
    grepl("^Rare CNS,", molecular_subtype),
    molecular_subtype,
    paste0("Rare CNS, ", molecular_subtype)
  )) %>%
  select(-OPC_molecular_subtype)


# make a methyl map
methyl_map <- rare_cns_methyl %>%
  select(match_id, molecular_subtype) %>%
  unique()

# any dups? no
length(unique(methyl_map$match_id)) == length(methyl_map$match_id)

# select all strategies for rare tumors
rare_hist <- hist %>%
  filter(match_id %in% methyl_map$match_id)

rare_cns_subtype_final <- rare_hist %>% 
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, match_id, sample_id, composition) %>%
  left_join(methyl_map) %>%
  # write result
  readr::write_tsv(file.path(results_dir, "rare-cns-molecular-subtypes.tsv"))


# how many tumors subtyped?
tumors_subtyped <- rare_cns_subtype_final %>%
  select(match_id, molecular_subtype) %>%
  unique() %>%
  count(molecular_subtype) %>%
  arrange(-n)

print(tumors_subtyped)

```

## Session Info

```{r}
sessionInfo()
```
