---
title: "Subset files for Rare CNS subtyping"
output: html_notebook
author: "Aylar Babaei, Jo Lynne Rokita"
date: 2026
---

In this PR, we identify candidate Rare CNS biospecimens from `histologies-base.tsv` and gather high-confidence methylation subtype calls for Rare CNS subtyping.

As per issue [#114](https://github.com/rokitalab/OpenPedCan-Project-CNH/issues/114), Rare CNS subtyping is based on high-confidence methylation subclassifications with `dkfz_v12_methylation_subclass_score >= 0.8` first, or NIH v2 subtypes with scores >= 0.9.

The following Rare CNS methylation classes are included in this step:

- Rare CNS, MN1::BEND2
- Rare CNS, ANTCON
- Rare CNS, ARMS
- Rare CNS, BCOR-fused
- Rare CNS, BCOR ITD
- Rare CNS, FOXR2 NBL
- Rare CNS, CIC SARC
- Rare CNS, DICER1 SARC
- Rare CNS, ERMS
- Rare CNS, BRD4::LEUTX
- Rare CNS, PLAG AMP
- Rare CNS, EWS
- Rare CNS, MN1::CXXC5
- Rare CNS, PATZ1-fused
- Rare CNS, PLAGL1-fused
- Rare CNS, MYOD1 RMS
- Rare CNS, CRINET

### Setup

```{r}
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

subset_dir <- file.path(
  root_dir,
  "analyses",
  "molecular-subtyping-Rare-CNS",
  "rare-cns-subset"
)

results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-Rare-CNS", "results")

if(!dir.exists(results_dir)){
  dir.create(results_dir)
}
```

## Input

```{r}
# File from 00-Rare-CNS-select-pathology-dx.R that is used for pathology
# diagnosis inclusion, exclusion criteria and Rare CNS methylation mapping
termslist <- jsonlite::fromJSON(
  file.path(subset_dir, "rare_cns_subtyping_path_dx_strings.json")
)

# subtype lookup table
subtype_to_opc_map <- tribble(
  ~NIH_v2_methylation_Class, ~dkfz_v12_methylation_subclass, ~OPC_molecular_subtype,
  "HGNET_MN1", "ABM_MN1", "Rare CNS, MN1::BEND2",
  "ANTCON", "ANTCON", "Rare CNS, ANTCON",
  "RMS_alevolar", "ARMS", "Rare CNS, ARMS",
  "EP300_BCOR", "CNS_BCOR_FUS", "Rare CNS, BCOR-fused",
  "HGNET_BCOR", "CNS_BCOR_ITD", "Rare CNS, BCOR ITD",
  "CNS_NB_FOXR2", "CNS_NB_FOXR2", "Rare CNS, FOXR2 NBL",
  "EFT_CIC", "CNS_SARC_CIC", "Rare CNS, CIC SARC",
  "CNS_SCS_DICER", "CNS_SARC_DICER", "Rare CNS, DICER1 SARC",
  "RMS_embryonal", "ERMS", "Rare CNS, ERMS",
  "ET_BRD4_LEUTX", "ET_BRD4_LEUTX", "Rare CNS, BRD4::LEUTX",
  "HGNET_PLAG", "ET_PLAG", "Rare CNS, PLAG AMP",
  "EWS", "EWS", "Rare CNS, EWS",
  "HGNET_CXXC5", "NET_CXXC5", "Rare CNS, MN1::CXXC5",
  "HGNET_PATZ", "NET_PATZ1", "Rare CNS, PATZ1-fused",
  "PLAGL1_FUS", "NET_PLAGL1_FUS", "Rare CNS, PLAGL1-fused",
  "RMS_MYOD1", "RMS_MYOD1", "Rare CNS, MYOD1 RMS",
  "CRINET", "CRINET", "Rare CNS, CRINET"
)

# Clinical file
hist <- readr::read_tsv(
  file.path(root_dir, "data", "histologies-base.tsv"),
  guess_max = 100000
)
```

## pathology-subset 

```{r}
# Candidate Rare CNS tumor specimens based on methylation
rare_cns_methyl <- hist %>%
  mutate(
    dkfz_pass = dkfz_v12_methylation_subclass_score >= 0.8 &
      dkfz_v12_methylation_subclass %in% termslist$include_dkfz_subtypes,
    nih_fallback_pass = (is.na(dkfz_v12_methylation_subclass_score) |
        dkfz_v12_methylation_subclass_score < 0.8) &
      NIH_v2_methylation_Superfamily_mean_score >= 0.9 &
      NIH_v2_methylation_Class_mean_score >= 0.9 &
      NIH_v2_methylation_Class %in% termslist$include_nih_subtypes
  ) %>%
  filter(experimental_strategy == "Methylation", dkfz_pass | nih_fallback_pass) %>%
  left_join(
    subtype_to_opc_map %>%
      transmute(dkfz_v12_methylation_subclass,
                dkfz_opc_molecular_subtype = OPC_molecular_subtype),
    by = "dkfz_v12_methylation_subclass"
  ) %>%
  left_join(
    subtype_to_opc_map %>%
      transmute(NIH_v2_methylation_Class,
                nih_opc_molecular_subtype = OPC_molecular_subtype),
    by = "NIH_v2_methylation_Class"
  ) %>%
  mutate(
    molecular_subtype = case_when(
      dkfz_pass ~ dkfz_opc_molecular_subtype,
      nih_fallback_pass ~ nih_opc_molecular_subtype,
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(molecular_subtype)) %>%
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, match_id, sample_id, composition, 
         dkfz_v12_methylation_subclass, dkfz_v12_methylation_subclass_score,
         NIH_v2_methylation_Class, NIH_v2_methylation_Superfamily_mean_score, NIH_v2_methylation_Class_mean_score, molecular_subtype)

# make a methyl mapping
methyl_map <- rare_cns_methyl %>%
  select(match_id, molecular_subtype) %>%
  unique()

# any dups? no
length(unique(methyl_map$match_id)) == length(methyl_map$match_id)

# select all strategies for rare tumors
rare_hist <- hist %>%
  filter(match_id %in% methyl_map$match_id)

rare_cns_subtype_final <- rare_hist %>% 
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, match_id, sample_id, composition) %>%
  left_join(methyl_map) %>%
  # write result
  readr::write_tsv(file.path(results_dir, "rare-cns-molecular-subtypes.tsv"))


# how many tumors subtyped?
tumors_subtyped <- rare_cns_subtype_final %>%
  select(match_id, molecular_subtype) %>%
  unique() %>%
  count(molecular_subtype) %>%
  arrange(-n)

print(tumors_subtyped)

```

## Session Info

```{r}
sessionInfo()
```
