---
title: "Subset files for Rare CNS subtyping"
output: html_notebook
author: "Aylar Babaei, Jo Lynne Rokita"
date: 2026
---

In this PR, we identify candidate Rare CNS biospecimens from `histologies-base.tsv` and gather high-confidence methylation subtype calls for Rare CNS subtyping.

As per issue [#114](https://github.com/rokitalab/OpenPedCan-Project-CNH/issues/114), Rare CNS subtyping is based on high-confidence methylation subclassifications with `dkfz_v12_methylation_subclass_score >= 0.8` first, or NIH v2 subtypes with scores >= 0.9. We additionally layer in a newer DKFZ v12.8 / NIH v3.1 classifier run (see `molecular-subtyping-methylation/input/bti_methylation_classifications.tsv`) using the same 0.8 / 0.9 thresholds and inclusion list, preferring the v12.8 call when a tumor has calls from both classifier versions.

The following Rare CNS methylation classes are included in this step:

- Rare CNS, MN1::BEND2
- Rare CNS, ANTCON
- Rare CNS, ARMS
- Rare CNS, BCOR-fused
- Rare CNS, BCOR ITD
- Rare CNS, FOXR2 NBL
- Rare CNS, CIC SARC
- Rare CNS, DICER1 SARC
- Rare CNS, ERMS
- Rare CNS, BRD4::LEUTX
- Rare CNS, PLAG AMP
- Rare CNS, EWS
- Rare CNS, MN1::CXXC5
- Rare CNS, PATZ1-fused
- Rare CNS, PLAGL1-fused
- Rare CNS, MYOD1 RMS
- Rare CNS, CRINET

### Setup

```{r}
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

subset_dir <- file.path(
  root_dir,
  "analyses",
  "molecular-subtyping-Rare-CNS",
  "rare-cns-subset"
)

results_dir <-
  file.path(root_dir, "analyses", "molecular-subtyping-Rare-CNS", "results")

if(!dir.exists(results_dir)){
  dir.create(results_dir)
}
```

## Input

```{r}
# File from 00-Rare-CNS-select-pathology-dx.R that is used for pathology
# diagnosis inclusion, exclusion criteria and Rare CNS methylation mapping
termslist <- jsonlite::fromJSON(
  file.path(subset_dir, "rare_cns_subtyping_path_dx_strings.json")
)

# subtype lookup table
subtype_to_opc_map <- tribble(
  ~NIH_v2_methylation_Class, ~dkfz_v12_methylation_subclass, ~OPC_molecular_subtype,
  "HGNET_MN1", "ABM_MN1", "Rare CNS, MN1::BEND2",
  "ANTCON", "ANTCON", "Rare CNS, ANTCON",
  "RMS_alevolar", "ARMS", "Rare CNS, ARMS",
  "EP300_BCOR", "CNS_BCOR_FUS", "Rare CNS, BCOR-fused",
  "HGNET_BCOR", "CNS_BCOR_ITD", "Rare CNS, BCOR ITD",
  "CNS_NB_FOXR2", "CNS_NB_FOXR2", "Rare CNS, FOXR2 NBL",
  "EFT_CIC", "CNS_SARC_CIC", "Rare CNS, CIC SARC",
  "CNS_SCS_DICER", "CNS_SARC_DICER", "Rare CNS, DICER1 SARC",
  "RMS_embryonal", "ERMS", "Rare CNS, ERMS",
  "ET_BRD4_LEUTX", "ET_BRD4_LEUTX", "Rare CNS, BRD4::LEUTX",
  "HGNET_PLAG", "ET_PLAG", "Rare CNS, PLAG AMP",
  "EWS", "EWS", "Rare CNS, EWS",
  "HGNET_CXXC5", "NET_CXXC5", "Rare CNS, MN1::CXXC5",
  "HGNET_PATZ", "NET_PATZ1", "Rare CNS, PATZ1-fused",
  "PLAGL1_FUS", "NET_PLAGL1_FUS", "Rare CNS, PLAGL1-fused",
  "RMS_MYOD1", "RMS_MYOD1", "Rare CNS, MYOD1 RMS",
  "CRINET", "CRINET", "Rare CNS, CRINET"
)

# Clinical file
hist <- readr::read_tsv(
  file.path(root_dir, "data", "histologies-base.tsv"),
  guess_max = 100000
)
```

## pathology-subset 

```{r}
# Candidate Rare CNS tumor specimens based on methylation
rare_cns_methyl <- hist %>%
  mutate(
    dkfz_pass = dkfz_v12_methylation_subclass_score >= 0.8 &
      dkfz_v12_methylation_subclass %in% termslist$include_dkfz_subtypes,
    nih_fallback_pass = (is.na(dkfz_v12_methylation_subclass_score) |
        dkfz_v12_methylation_subclass_score < 0.8) &
      NIH_v2_methylation_Superfamily_mean_score >= 0.9 &
      NIH_v2_methylation_Class_mean_score >= 0.9 &
      NIH_v2_methylation_Class %in% termslist$include_nih_subtypes
  ) %>%
  filter(cohort == "PBTA", experimental_strategy == "Methylation", dkfz_pass | nih_fallback_pass) %>%
  left_join(
    subtype_to_opc_map %>%
      transmute(dkfz_v12_methylation_subclass,
                dkfz_opc_molecular_subtype = OPC_molecular_subtype),
    by = "dkfz_v12_methylation_subclass"
  ) %>%
  left_join(
    subtype_to_opc_map %>%
      transmute(NIH_v2_methylation_Class,
                nih_opc_molecular_subtype = OPC_molecular_subtype),
    by = "NIH_v2_methylation_Class"
  ) %>%
  mutate(
    molecular_subtype = case_when(
      dkfz_pass ~ dkfz_opc_molecular_subtype,
      nih_fallback_pass ~ nih_opc_molecular_subtype,
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(molecular_subtype)) %>%
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, match_id, sample_id, composition, 
         dkfz_v12_methylation_subclass, dkfz_v12_methylation_subclass_score,
         NIH_v2_methylation_Class, NIH_v2_methylation_Superfamily_mean_score, NIH_v2_methylation_Class_mean_score, pathology_diagnosis, molecular_subtype) %>%
  # remove samples with pathology diagnosis of MB with Rare CNS, ARMS subtype --> these can mis-cluster if they are myogenic MB
  filter(!(pathology_diagnosis == "Medulloblastoma" & molecular_subtype == "Rare CNS, ARMS"))

# make a methyl mapping from the histologies-base.tsv (dkfz v12 / NIH v2) calls above
methyl_map_v12 <- rare_cns_methyl %>%
  select(match_id, molecular_subtype) %>%
  unique()
```

## New DKFZ v12.8 methylation classifications

A newer classifier run (DKFZ v12.8 / NIH v3.1) was dropped in
`../molecular-subtyping-methylation/input/bti_methylation_classifications.tsv`,
keyed by `Bioassay_ID`. Per the source, a `BS_`-prefixed `Bioassay_ID` is a
`Kids_First_Biospecimen_ID`; the original file also included `BA_`-prefixed
IDs (external/non-PBTA samples), which have already been filtered out of this
subset. Any `BS_` ID not found in this cohort is still dropped here via the
join below. DKFZ v12.8
subclass abbreviations are largely the same taxonomy as v12, but not always:
DKFZ renamed `ANTCON` to `GTAKA` in v12.8 (confirmed on specimens with a
near-1.0-confidence call under both versions), so we alias it back before
matching against the v12 inclusion list / subtype map.

NIH v3.1 uses a different abbreviation scheme than NIH v2 for its fine-grained
call (e.g. `CNS BCOR ITD` vs. `HGNET_BCOR`, and `GTAKA` again for `ANTCON`) --
note v3.1's `subclass` column, not `class`, is the analog of v2's `Class`
column used for matching. We derived the alias table below empirically, by
pairing specimens with a confirmed high-confidence DKFZ v12 call against their
NIH v3.1 subclass call. `ET_BRD4_LEUTX`, `ERMS`, `RMS_MYOD1`, and `CRINET`
have no high-confidence specimens in the cohort yet to confirm their v3.1
code, so they're intentionally left out of this table (and out of the NIH
v3.1 fallback) until one is observed.

```{r}
bti_file <- file.path(
  root_dir, "analyses", "molecular-subtyping-methylation",
  "input", "bti_methylation_classifications.tsv"
)

# known DKFZ subclass renames between v12 and v12.8
dkfz_v128_to_v12_alias <- c("GTAKA" = "ANTCON")

# empirically-derived NIH v3.1 subclass -> DKFZ v12 subclass vocabulary,
# so v3.1 calls can reuse termslist$include_dkfz_subtypes / subtype_to_opc_map
nih_v31_subclass_alias <- c(
  "EWS" = "EWS",
  "CNS BCOR ITD" = "CNS_BCOR_ITD",
  "CNS SARC CIC" = "CNS_SARC_CIC",
  "CNS PLAGL1 FUS" = "NET_PLAGL1_FUS",
  "GTAKA" = "ANTCON",
  "RMS ALV" = "ARMS",
  "CNS PATZ1 FUS" = "NET_PATZ1",
  "AB MN1 BEND2" = "ABM_MN1",
  "CNS NB FOXR2" = "CNS_NB_FOXR2",
  "CNS PLAG AMP" = "ET_PLAG",
  "CNS BCOR FUS" = "CNS_BCOR_FUS",
  "CNS SARC DICER" = "CNS_SARC_DICER",
  "AB MN1 CXXC5" = "NET_CXXC5"
)

bti_calls <- readr::read_tsv(bti_file, guess_max = 10000) %>%
  rename(Kids_First_Biospecimen_ID = Bioassay_ID) %>%
  mutate(
    dkfz_v12.8_subclass = recode(dkfz_v12.8_subclass, !!!dkfz_v128_to_v12_alias),
    NIH_v3.1_subclass = recode(NIH_v3.1_subclass, !!!nih_v31_subclass_alias),
    dkfz_v12.8_pass = dkfz_v12.8_subclass_maxscore >= 0.8 &
      dkfz_v12.8_subclass %in% termslist$include_dkfz_subtypes,
    nih_v31_pass = (is.na(dkfz_v12.8_subclass_maxscore) | dkfz_v12.8_subclass_maxscore < 0.8) &
      NIH_v3.1_superfamily_meanscore >= 0.9 &
      NIH_v3.1_class_score >= 0.9 &
      NIH_v3.1_subclass %in% termslist$include_dkfz_subtypes
  ) %>%
  filter(dkfz_v12.8_pass | nih_v31_pass) %>%
  left_join(
    subtype_to_opc_map %>%
      transmute(dkfz_v12.8_subclass = dkfz_v12_methylation_subclass,
                dkfz_opc_molecular_subtype = OPC_molecular_subtype),
    by = "dkfz_v12.8_subclass"
  ) %>%
  left_join(
    subtype_to_opc_map %>%
      transmute(NIH_v3.1_subclass = dkfz_v12_methylation_subclass,
                nih_opc_molecular_subtype = OPC_molecular_subtype),
    by = "NIH_v3.1_subclass"
  ) %>%
  mutate(
    molecular_subtype = case_when(
      dkfz_v12.8_pass ~ dkfz_opc_molecular_subtype,
      nih_v31_pass ~ nih_opc_molecular_subtype,
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(molecular_subtype)) %>%
  # keep only specimens that are part of this cohort
  inner_join(
    hist %>% filter(cohort == "PBTA") %>%
      select(Kids_First_Biospecimen_ID, match_id, pathology_diagnosis),
    by = "Kids_First_Biospecimen_ID"
  ) %>%
  # same mis-clustering safeguard as the v12/v2 calls above
  filter(!(pathology_diagnosis == "Medulloblastoma" & molecular_subtype == "Rare CNS, ARMS")) %>%
  distinct(match_id, molecular_subtype)

nrow(bti_calls)
```

Combine the v12/v2-based calls with the new v12.8-based calls. Where a
`match_id` has a call from both, prefer the newer v12.8 classifier, but flag
any disagreement for manual review rather than silently overwriting.

```{r}
combined_calls <- full_join(
  methyl_map_v12 %>% rename(molecular_subtype_v12 = molecular_subtype),
  bti_calls %>% rename(molecular_subtype_v128 = molecular_subtype),
  by = "match_id"
)

# tumors where both classifiers ran and disagree on subtype
methyl_conflicts <- combined_calls %>%
  filter(!is.na(molecular_subtype_v12), !is.na(molecular_subtype_v128),
         molecular_subtype_v12 != molecular_subtype_v128)

print(methyl_conflicts)

methyl_map <- combined_calls %>%
  mutate(molecular_subtype = coalesce(molecular_subtype_v128, molecular_subtype_v12)) %>%
  select(match_id, molecular_subtype)

# any dups? no
length(unique(methyl_map$match_id)) == nrow(methyl_map)
```

```{r}
# select all strategies for rare tumors
rare_hist <- hist %>%
  filter(match_id %in% methyl_map$match_id)

rare_cns_subtype_final <- rare_hist %>% 
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, match_id, sample_id, composition) %>%
  left_join(methyl_map) %>%
  # write result
  readr::write_tsv(file.path(results_dir, "rare-cns-molecular-subtypes.tsv"))


# how many tumors subtyped?
tumors_subtyped <- rare_cns_subtype_final %>%
  select(match_id, molecular_subtype) %>%
  unique() %>%
  count(molecular_subtype) %>%
  arrange(-n)

print(tumors_subtyped)

```

## Confirm fusion evidence agrees with the methylation-based subtype

Fusions (`01-subtype-using-fusions.Rmd`) are not used to add new samples or subtypes to this cohort. For tumors that already have a high-confidence methylation subtype above and RNA-seq available, we check that any hallmark fusion detected agrees with the methylation-based call, and flag any that don't for manual review.

```{r}
fusion_results_file <- file.path(results_dir, "rare_cns_fusion_based_subtypes.tsv")

fusion_map <- readr::read_tsv(fusion_results_file, col_types = cols(.default = "c")) %>%
  filter(!is.na(molecular_subtype)) %>%
  distinct(match_id, molecular_subtype) %>%
  group_by(match_id) %>%
  summarise(fusion_molecular_subtype = paste(sort(unique(molecular_subtype)), collapse = ";"),
            .groups = "drop")

# tumors already classified above that have RNA-seq, and so could have fusion evidence
rna_available <- hist %>%
  filter(match_id %in% rare_cns_subtype_final$match_id,
         !is.na(RNA_library)) %>%
  distinct(match_id)

pathology_dx_by_match_id <- hist %>%
  filter(match_id %in% rare_cns_subtype_final$match_id) %>%
  distinct(match_id, pathology_diagnosis)

fusion_methyl_confirmation <- rare_cns_subtype_final %>%
  distinct(match_id, molecular_subtype) %>%
  inner_join(rna_available, by = "match_id") %>%
  left_join(fusion_map, by = "match_id") %>%
  left_join(pathology_dx_by_match_id, by = "match_id") %>%
  mutate(concordant = case_when(
    is.na(fusion_molecular_subtype) ~ NA,
    fusion_molecular_subtype == molecular_subtype ~ TRUE,
    TRUE ~ FALSE
  )) %>%
  write_tsv(file.path(results_dir, "rare_cns_fusion_methylation_confirmation.tsv"))

# tumors with RNA-seq and a fusion call that disagrees with the methylation-based subtype
# note: a BCOR ITD methylation call with BCOR-fused fusion evidence is not
# necessarily an error -- ITD (internal tandem duplication) and a fusion are
# distinct events that can co-occur; still flagged here for manual review.
discordant <- fusion_methyl_confirmation %>%
  filter(concordant == FALSE)

print(discordant)
```

## Verify one subtype per biospecimen

```{r}
multi_subtype_bs_ids <- rare_cns_subtype_final %>%
  distinct(Kids_First_Biospecimen_ID, molecular_subtype) %>%
  count(Kids_First_Biospecimen_ID) %>%
  filter(n > 1)

if (nrow(multi_subtype_bs_ids) > 0) {
  stop("The following Kids_First_Biospecimen_ID(s) have more than one Rare CNS molecular_subtype assigned: ",
       paste(multi_subtype_bs_ids$Kids_First_Biospecimen_ID, collapse = ", "))
}
```

## Session Info

```{r}
sessionInfo()
```
